\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\begin{document}

\title{Notes on the Trotter Breakup for Jaynes-Cummings Hamiltonians}
\author{Matthew Otten}

\maketitle

\begin{abstract}
  Basic notes on utilizing the trotter breakup for the Jaynes-Cummings Hamiltonian.
\end{abstract}

\section{Hamiltonian}
The Jaynes-Cummings model describeds the dynamics of
a two level system coupled to an oscillator mode (be it a mechanical resonator, a
plasmonic system, or an electromagnetic field):
\begin{equation}\label{jc_ham}
  H =  \omega_a a^\dagger a +  \omega_\sigma \sigma^\dagger \sigma +
   g (\sigma + \sigma^\dagger ) ( a + a^\dagger),
\end{equation}
where $a^\dagger$ is the creation operator of the oscillator, $\omega_s$ is the frequency
of the oscillator, $\sigma^\dagger$ is the creation operator of the two level system,
$\omega_\sigma$ is the transition frequency of the two level sistem, and $g$ is the
coupling strength between the two systems. We are using units such that $\hbar = 1$.
This Hamiltonian can be extended to many two level systems, which is sometimes
called the Tavis-Cummings Hamiltonian:
\begin{equation}\label{tc_ham}
  H =  \omega_a a^\dagger a +  \sum_i^{N_{tls}}\omega_i \sigma_i^\dagger \sigma_i +
   g_i ( \sigma_i + \sigma_i^\dagger) ( a + a^\dagger),
\end{equation}
which is just equation~\ref{jc_ham}, but with multiple two level systems,
$\sigma_i$. Using either of these Hamiltonians, we can solve for the time
dynamics of the system using the time-dependent Schrodinger equation,

\begin{equation}\label{schrod}
  \dot{\psi} = -i H \psi,
\end{equation}
which has the solution

\begin{equation}\label{solution}
  \psi (t) = \exp(-i H t) \psi (0).
\end{equation}
If the exponential of $H$ can be calculated efficiently, then the problem is
solved.

\section{Trotter Breakup}
Taking the exponential of a matrix is, generally, difficult. One avenue is using
the symmetric Trotter breakup to split $H$ into two parts: the diagonal
oscillator terms ($A = \omega_a a^\dagger a + \sum_i^{N_{tls}}\omega_i
\sigma_i^\dagger \sigma_i$) and the off diagonal coupling terms ($B =
\sum_i^{N_{tls}} g_i ( \sigma_i \sigma_i^\dagger) ( a + a^\dagger),$). The
symmetric Trotter breakup states that
\begin{equation} \label{trotter}
  \exp(-i(A+B)\Delta t) \approx \exp(-i A \Delta t/2) \exp(-i B \Delta t) \exp(-i A \Delta t/2) + \mathcal{O} (\Delta t ^3).
\end{equation}
The exponential of a diagonal matrix, such as $A$, is trivially the exponential
of the diagonal elements. If the exponential of $B$ can be found with similar
ease, the explicit propagator can be constructed and time stepping becomes a
matrix vector product with $2^{nd}$ order accuracy. One way of finding the
exponential of a matrix is by diagonalizing that matrix.

\section{Eigenvectors of Kronecker Products}
Luckily, $B$ is a matrix of high structure. In the one two level system case,
explicitly including the tensor products
\begin{equation}\label{explicit_b}
  B = (a + a^\dagger) \otimes (\sigma + \sigma^\dagger ).
\end{equation}
In the multiple two level system case, there are additional Kronecker products.
The eigenvectors of the Kronecker product of two matrices turns
out to be the Kronecker product of the eigenvectors of the two matrices, and the
eigenvalues of the Kronecker product are the product of the eigenvalues of the
two matrices. This can be simply shown,
\begin{equation}\label{kron_ev}
  (C\otimes D) (v_c \otimes v_d) = (C v_c ) \otimes (D v_d) = (\lambda_c v_c) \otimes (\lambda_d v_d) = \lambda_c \lambda_d (v_c \otimes v_d).
\end{equation}

Since our $B$ deals with the tensor product of many small matrices,
equation~\ref{kron_ev} shows that we can build the eigensystems
of huge matrices ($2^{30}$) using the eigensystems of the small matrices ($<\mathcal{O}
(10)$). One issue is that if $C = D$, the eigenvectors obtained via
equation~\ref{kron_ev} are actually a linear combination of the basic
eigenvectors (at least in the case where $C = \sigma + \sigma^\dagger$). I am
not sure exactly how to deal with this in the general case, but I found an
empirical way of getting the eigenvectors of $C \otimes C \otimes C \otimes ...
\otimes C$ when $C = \sigma + \sigma^\dagger$, which is good enough for now.
Then, by construction, I can get the eigenvectors and eigenvalues of $B$
(without any need to do explicit diagonalization). Using this, and the Trotter
breakup of equation~\ref{trotter}, we can create the explicit propagator:
\begin{equation}\label{propagator}
  P = \exp(-i A \Delta t/2) U \exp(-i V \Delta t) U^\dagger \exp(-i A \Delta t/2),
\end{equation}
where $U$ is the matrix of eigenvectors of $B$ and $V$ is the diagonal matrix of
eigenvalues of $B$. With $P$, we can do the time propagation efficiently - each
timestep is a single matrix - vector product, and the method is stable with any
timestep, with second order accuracy.

\end{document}
